An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-07-12 paper/main.tex
formatting

% Packages
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{float}
\usepackage{derivative}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}

@@

	Artificial Neural Networks (ANNs) are a key part of today’s machine learning systems. They’ve made huge progress in areas like image recognition, speech processing, and self-driving cars. However, one of the biggest challenges with ANNs is that they use a lot of memory and computing power. This becomes a problem when you want to run them on devices with limited resources, such as smartphones, embedded systems, or edge devices. \\ \\
	To solve this, researchers have looked into ways to make neural networks more efficient. A common approach is to reduce the size of the network or shrink the input batch size. But doing this can harm the network’s ability to learn and generalize. A better strategy is to reduce the numerical precision of the network’s parameters such as things like weights, activations, and gradients. This method, called quantization, makes the model smaller and faster while using less energy, without necessarily hurting accuracy. \\ \\
	One advanced version of this method is called mixed-precision quantization. Instead of using the same number of bits for every layer in the network, mixed-precision assigns different bitwidths depending on how important each part is. For example, the early layers that handle low-level features might use higher precision, while deeper layers could work fine with fewer bits. This flexible approach often leads to better results than using a single fixed precision for the whole model. Researchers have used a range of techniques, like gradient descent, heuristics, and even evolutionary algorithms, to find the best bitwidth settings for each layer. \\ \\
	Another method that helps reduce resource usage is network pruning. This means cutting out parts of the network that don’t contribute much to its performance. One interesting way to do this is with simulated annealing, a search algorithm that mimics the process of cooling metal to find a stable structure. It doesn’t need gradient information and can still find good network shapes by exploring many possibilities. This makes it useful when you want to shrink a network without going through heavy retraining. \\ \\
	Previous studies have shown that reducing the precision of neural network parameters can lead to significant improvements in memory usage and processing speed. Even when precision is reduced to very low levels, models can still perform competitively on complex tasks. However, while these early results are promising, the overall effects of lowering precision, especially how it influences model accuracy, training behavior, and storage efficiency, are still not fully understood and deserve more detailed investigation. \\ \\
	This is where our experimental work begins. 
	In this paper, we take a closer look at how different levels of parameter precision affect the behavior and performance of neural networks. Our goal is to understand what happens when we reduce precision in various ways, and how much we can lower it before performance starts to drop. \\ \\
	We start by testing how reducing precision during both training and inference affects model accuracy and runtime. Next, we look at what happens when we train the network using full precision, but lower the precision only during inference. This setup is common in real-world applications, where a powerful machine handles training, but the final model is deployed on a lightweight device. 
	Then, we run a series of tests where we round the trained weights to a limited number of decimal places. This helps us see how much we can simplify the stored weight values without hurting accuracy too much. This is especially useful for devices with very limited storage. 
	We also try out mixed-precision training, using a mix of FP16 and FP32 formats. On top of that, we experiment with different types of gradient clipping, like global norm clipping and value clipping, to see how they affect training stability and final accuracy. 
	Even though we use relatively simple network architectures and datasets, our results show that neural networks are quite robust to low-precision settings. In some cases, we were able to round weights to just three decimal digits and still get nearly the same accuracy as the full-precision version. 
	These findings show that it’s possible to significantly reduce the memory and compute needs of neural networks without losing much performance. This makes them more practical for use in low-resource settings. It also opens the door for more advanced techniques that combine quantization, pruning, and precision-aware training to build highly efficient models for real-world applications.

@@

  	\begin{align*}
    	L(f(x_1, x_2)) &= (\sigma(\omega^T x + b) - y) ^2\\
                   	&= (\sigma(a) - y) ^ 2\\
                   	&= (\hat y - y) ^ 2
  	\end{align*}
  	\begin{alignat*}{3}
    	\pdv{L}{\omega_1} &= \pdv{L}{\sigma} &&\pdv{\sigma}{a}      &&\pdv{a}{\omega_1}\\
                      	&= 2(\hat y - y)~~ &&\odv{}{a}\sigma(a)~~ &&x_1
  	\end{alignat*}

@@

	\subsection*{Quantization During Training}
	Several studies have explored training neural networks with reduced precision directly. Notable contributions include:
	\begin{itemize}
    	\item \textbf{Mixed Precision Training}, which leverages float16 arithmetic for speed while maintaining float32 master weights \cite{micikevicius2017mixed}.
    	\item \textbf{DoReFa-Net}, which quantizes weights, activations, and gradients to arbitrary bitwidths \cite{zhou2018dorefa}.
    	\item \textbf{Q-BERT}, a quantized BERT model that maintains accuracy using only 4-bit weights \cite{shen2020qbert}.
	\end{itemize}

	\subsection*{Post-Training Quantization}
	Other approaches focus on converting pretrained models into lower-precision equivalents:
	\begin{itemize}
    	\item \textbf{Integer-Only Quantization} using symmetric quantization scales \cite{jacob2018quantization}.
    	\item \textbf{GPTQ}, a method for quantizing large transformer models without retraining \cite{frantar2022gptq}.
    	\item \textbf{SmoothQuant}, which scales activations to better match weight ranges during quantization \cite{xiao2024smoothquant}.
	\end{itemize}

@@

	\begin{figure}[H]
    	\centering
    	\includegraphics[width=0.75\textwidth]{figures/118_accuracy.png}
    	\caption{Accuracy vs. number of decimal digits retained in weights.}
	\end{figure}

@@

	\newpage
	\printbibliography
\end{document}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{float}
\usepackage{derivative}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}

@@

Artificial Neural Networks (ANNs) are a key part of today’s machine learning
systems. They’ve made huge progress in areas like image recognition, speech
processing, and self-driving cars. However, one of the biggest challenges with
ANNs is that they use a lot of memory and computing power. This becomes a
problem when you want to run them on devices with limited resources, such as
smartphones, embedded systems, or edge devices. \\ \\ To solve this,
researchers have looked into ways to make neural networks more efficient. A
common approach is to reduce the size of the network or shrink the input batch
size. But doing this can harm the network’s ability to learn and generalize. A
better strategy is to reduce the numerical precision of the network’s
parameters such as things like weights, activations, and gradients. This
method, called quantization, makes the model smaller and faster while using
less energy, without necessarily hurting accuracy. \\ \\ One advanced version
of this method is called mixed-precision quantization. Instead of using the
same number of bits for every layer in the network, mixed-precision assigns
different bitwidths depending on how important each part is. For example, the
early layers that handle low-level features might use higher precision, while
deeper layers could work fine with fewer bits. This flexible approach often
leads to better results than using a single fixed precision for the whole
model. Researchers have used a range of techniques, like gradient descent,
heuristics, and even evolutionary algorithms, to find the best bitwidth
settings for each layer. \\ \\ Another method that helps reduce resource usage
is network pruning. This means cutting out parts of the network that don’t
contribute much to its performance. One interesting way to do this is with
simulated annealing, a search algorithm that mimics the process of cooling
metal to find a stable structure. It doesn’t need gradient information and can
still find good network shapes by exploring many possibilities. This makes it
useful when you want to shrink a network without going through heavy
retraining. \\ \\ Previous studies have shown that reducing the precision of
neural network parameters can lead to significant improvements in memory usage
and processing speed. Even when precision is reduced to very low levels, models
can still perform competitively on complex tasks. However, while these early
results are promising, the overall effects of lowering precision, especially
how it influences model accuracy, training behavior, and storage efficiency,
are still not fully understood and deserve more detailed investigation. \\ \\
This is where our experimental work begins. In this paper, we take a closer
look at how different levels of parameter precision affect the behavior and
performance of neural networks. Our goal is to understand what happens when we
reduce precision in various ways, and how much we can lower it before
performance starts to drop. \\ \\ We start by testing how reducing precision
during both training and inference affects model accuracy and runtime. Next, we
look at what happens when we train the network using full precision, but lower
the precision only during inference. This setup is common in real-world
applications, where a powerful machine handles training, but the final model is
deployed on a lightweight device. Then, we run a series of tests where we round
the trained weights to a limited number of decimal places. This helps us see
how much we can simplify the stored weight values without hurting accuracy too
much. This is especially useful for devices with very limited storage. We also
try out mixed-precision training, using a mix of FP16 and FP32 formats. On top
of that, we experiment with different types of gradient clipping, like global
norm clipping and value clipping, to see how they affect training stability and
final accuracy. Even though we use relatively simple network architectures and
datasets, our results show that neural networks are quite robust to
low-precision settings. In some cases, we were able to round weights to just
three decimal digits and still get nearly the same accuracy as the
full-precision version. These findings show that it’s possible to significantly
reduce the memory and compute needs of neural networks without losing much
performance. This makes them more practical for use in low-resource settings.
It also opens the door for more advanced techniques that combine quantization,
pruning, and precision-aware training to build highly efficient models for
real-world applications.

@@

\begin{align*}
	L(f(x_1, x_2)) & = {(\sigma(\omega^T x + b) - y)}^2 \\
	               & = {(\sigma(a) - y)}^2              \\
	               & = {(\hat y - y)}^2
\end{align*}
\begin{alignat*}{3}
	\pdv{L}{\omega_1} & = \pdv{L}{\sigma} &  & \pdv{\sigma}{a}      &  & \pdv{a}{\omega_1} \\
	                  & = 2(\hat y - y)~~ &  & \odv{}{a}\sigma(a)~~ &  & x_1
\end{alignat*}

@@

\subsection*{Quantization During Training}
Several studies have explored training neural networks with reduced precision
directly. Notable contributions include:
\begin{itemize}
	\item \textbf{Mixed Precision Training}, which leverages float16 arithmetic for speed while maintaining float32 master weights~\cite{micikevicius2017mixed}.
	\item \textbf{DoReFa-Net}, which quantizes weights, activations, and gradients to arbitrary bitwidths~\cite{zhou2018dorefa}.
	\item \textbf{Q-BERT}, a quantized BERT model that maintains accuracy using only 4-bit weights~\cite{shen2020qbert}.
\end{itemize}

\subsection*{Post-Training Quantization}
Other approaches focus on converting pretrained models into lower-precision
equivalents:
\begin{itemize}
	\item \textbf{Integer-Only Quantization} using symmetric quantization scales~\cite{jacob2018quantization}.
	\item \textbf{GPTQ}, a method for quantizing large transformer models without retraining~\cite{frantar2022gptq}.
	\item \textbf{SmoothQuant}, which scales activations to better match weight ranges during quantization~\cite{xiao2024smoothquant}.
\end{itemize}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\textwidth]{figures/118_accuracy.png}
	\caption{Accuracy vs.\ number of decimal digits retained in weights.}
\end{figure}

@@

\newpage
\printbibliography%
\end{document}
